Manuscript Template for a Data Analysis Project

### 2023-04-08 code/analysis_code/Race/p1_EDA_race.Rmd
added more to project

```{r}
autoplot(lm(race_df$`Death Rate (AIDSVu)` ~ 1))
autoplot(lm(race_df$`Death Rate (estimated)` ~ 1))
autoplot(lm(race_df$`Ratio Observed/Expected` ~ 1))
```

@@

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/Expected`, fill = Race), bins = 10)
shapiro.test(race_df$`Ratio Observed/Expected`)
```
Both histograms show a positively skewed distribution. ROE appears to be the slightest bit more normal.

@@

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Ratio Observed/Expected` + 1), fill = Race) , bins = 10)
shapiro.test(log(race_df$`Ratio Observed/Expected` + 1))
```
The log transformation of Death Rate better fit, but there is still sufficient evidence to reject normality.

@@

race_df_ROE <- race_df %>% 
  filter(`Ratio Observed/Expected` > 0)

b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/Expected` ~ 1))
lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]
```

@@

Box Cox transformation of ROE and results of Shapiro test.
```{r}
race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)

race_df_ROE <- race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE))

race_df_ROE %>%
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Black") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "White") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Hispanic") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Multiple") %>% 
  pull(BOX) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(race_df_ROE$BOX ~ 1))
```

Let's save the dataframe for ROE since after the Box Cox transformation for all races was better with ROE than DR.
```{r}
saveRDS(race_df_ROE, here("data/processed_data/race_new.rds"))
```

>>>>

```{r}
autoplot(lm(race_df$`Death Rate (AIDSVu)` ~ 1))
autoplot(lm(race_df$`Death Rate (estimated)` ~ 1))
autoplot(lm(race_df$`Ratio Observed/ Expected` ~ 1))
```

@@

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/ Expected`, fill = Race), bins = 10)
shapiro.test(race_df$`Ratio Observed/ Expected`)
```
Both histograms show a positively skewed distribution.

@@

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Ratio Observed/ Expected` + 1), fill = Race) , bins = 10)
shapiro.test(log(race_df$`Ratio Observed/ Expected` + 1))
```
The log transformation of Death Rate better fit, but there is still sufficient evidence to reject normality.

@@

race_df_ROE <- race_df %>% 
  filter(`Ratio Observed/ Expected` > 0)

b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/ Expected` ~ 1))
lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]
lambda_ROE 
```

@@

Box Cox transformation of ROE and results of Shapiro test.
```{r}
race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/ Expected`, lambda_ROE)) %>%
  ggplot() +
  geom_histogram(aes(x= BOX, fill = Race), bins = 10) +
  facet_wrap(. ~ Race) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Box Cox Transformation of Observed/ Expected HIV Mortality Ratio", y = "Count")

ggsave(here("results/box_hist.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
race_df_ROE <- race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/ Expected`, lambda_ROE))

race_df_ROE %>%
  pull(BOX) %>% 
  shapiro.test()
## Fail to reject normality

race_df_ROE %>%
  filter(Race == "Black") %>% 
  pull(BOX) %>% 
  shapiro.test()
## Fail to reject normality

race_df_ROE %>%
  filter(Race == "White") %>% 
  pull(BOX) %>% 
  shapiro.test()
## Reject normality

race_df_ROE %>%
  filter(Race == "Hispanic") %>% 
  pull(BOX) %>% 
  shapiro.test()
## Fail to reject normality

race_df_ROE %>%
  filter(Race == "Multiple") %>% 
  pull(BOX) %>% 
  shapiro.test() 
## Fail to reject normality

par(mfrow=c(1,3))
plot(lm(race_df_ROE$BOX ~ 1))
##I saved the plot above manually because I did not know how to save all three plots in one image aside from exporting.
```

The BOX transformation of ROE was better than that of DR, so let's use that one... but before wrapping up, let's go back and pull the observations that were dropped from the Box Cox transformation. 
```{r}
race_drop <- race_df %>% 
  filter(`Ratio Observed/ Expected` == 0)

race_new <- full_join(race_df_ROE, race_drop)

##We need to replace the NA with a value in order to run spatial tests later. We will replace with 0. 
race_new <-  race_new %>%
  mutate(BOX = replace(BOX, is.na(BOX),0))
```

```{r}
saveRDS(race_new, here("data/processed_data/race_new.rds"))
```

### 2023-04-08 code/analysis_code/Race/p2_PCA_race.Rmd
revisions based on feedback
added more to project

```{r}
pca_function <- function(race) {   
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

num_race <- df[,c(9:10, 16:25)]

@@

# Biplot of the attributes
print(fviz_pca_var(pca_race, col.var = "cos2",
             gradient.cols = c("darkblue", "orange", "green"),
             repel = TRUE))
# Variables in green are a good representation of the variable on that component.
}
```

```{r}
pca_function("Black")
```

```{r}
pca_function("Hispanic")
```

```{r}
pca_function("Multiple")
```

```{r}
pca_function("White")
```

>>>>

CREATING PCA FUNCTION
```{r}
pca_function <- function(race) {   
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

#Selecting variables
num_race <- df[,c(9:10, 16:25)]

@@

# Biplot of the attributes
print(fviz_pca_var(pca_race, col.var = "cos2",
             gradient.cols = c("darkblue", "orange", "green"),
             repel = TRUE, title = race))
# Variables in green are a good representation of the variable on that component.
}
```

```{r}
pca_function("Black")
ggsave(here("results/pca_black.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
pca_function("Hispanic")
ggsave(here("results/pca_hispanic.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
pca_function("Multiple")
ggsave(here("results/pca_multiple.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
pca_function("White")
ggsave(here("results/pca_white.png"), units="in", width=10, height=7, dpi=300)
```

### 2023-04-08 code/analysis_code/Race/p3_model_race.Rmd
revisions based on feedback
added more to project

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(spdep)
```

@@

  car_race <- spautolm(log_rate ~ `Percent Living in Poverty` + 
                             `Percent High School Education`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))
}
```

@@

@@

@@

@@

Fitting Negative binomial models
```{r}
nb_function("Black")
```

```{r}
nb_function("Hispanic")
```

@@

```{r}
nb_function("White")
```
The NB model accounting for overdispersion fit better than the poisson model.

@@

@@

```{r}
car_function("Multiple")
```

```{r}
car_function("White")
```
The model accounting for spatial variation provide the best fit.

@@

 # plot Moran's I
  par(mfrow=c(1,2))
  print(moran.test(df$BOX, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
  moran.plot(df$BOX, nablist_race_df, zero.policy = TRUE)

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$`Ratio Observed/Expected`, nablist_race_df)
  summary(lisa_race[,1])
}
```

@@

@@

@@

@@

```{r}
save.image(file="../../data/processed_data/race_env.rds")
```

>>>>

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(MASS)
library(spatialreg)
library(spdep)
```

@@

  car_race <- spautolm(`Death Cases` ~ `Median Household Income` +
                              `Percent High School Education` + 
                              offset(log(`Population`)),
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))
}
```

@@

```{r}
sink(here("results/pois_summary_black.txt"))
pois_function("Black")
sink()
```

@@

```{r}
sink(here("results/pois_summary_hispanic.txt"))
pois_function("Hispanic")
sink()
```

@@

```{r}
sink(here("results/pois_summary_multiple.txt"))
pois_function("Multiple")
sink()
```

@@

```{r}
sink(here("results/pois_summary_white.txt"))
pois_function("White")
sink()
```

Fitting Negative binomial models
```{r}
nb_function("Black")
```
```{r}
sink(here("results/nb_summary_black.txt"))
nb_function("Black")
sink()
```

```{r}
nb_function("Hispanic")
```
```{r}
sink(here("results/nb_summary_hispanic.txt"))
nb_function("Hispanic")
sink()
```

@@

```{r}
sink(here("results/nb_summary_multiple.txt"))
nb_function("Multiple")
sink()
```

```{r}
nb_function("White")
```

```{r}
sink(here("results/nb_summary_white.txt"))
nb_function("White")
sink()
```

The NB model accounting for overdispersion fits better than the poisson model.

@@

```{r}
sink(here("results/car_summary_black.txt"))
car_function("Black")
sink()
```

@@

```{r}
sink(here("results/car_summary_hispanic.txt"))
car_function("Hispanic")
sink()
```

```{r}
car_function("Multiple")
```
```{r}
sink(here("results/car_summary_multiple.txt"))
car_function("Multiple")
sink()
```

```{r}
car_function("White")
```
```{r}
sink(here("results/car_summary_white.txt"))
car_function("White")
sink()
```
The model accounting for spatial variation provide the best fit.

@@

 # plot Moran's I
  print(moran.test(df$BOX, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
  moran.plot(df$BOX, nablist_race_df, zero.policy = TRUE)

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$`Ratio Observed/ Expected`, nablist_race_df)
  summary(lisa_race[,1])
}
```

@@

```{r}
png(here("results/moran_black.png"))
sink(here("results/spacy_summary_black.txt"))
spacy_function("Black")
sink()
dev.off()
```

@@

```{r}
png(here("results/moran_hispanic.png"))
sink(here("results/spacy_summary_hispanic.txt"))
spacy_function("Hispanic")
sink()
dev.off()
```

@@

```{r}
png(here("results/moran_multiple.png"))
sink(here("results/spacy_summary_multiple.txt"))
spacy_function("Multiple")
sink()
dev.off()
```

@@

```{r}
png(here("results/moran_white.png"))
sink(here("results/spacy_summary_white.txt"))
spacy_function("White")
sink()
dev.off()
```

```{r}
save.image(here("data/processed_data/race_env.rds"))
```

### 2023-04-08 code/analysis_code/archieved/Sex/1_EDA_sex.Rmd
added more to project

Visual assessment and the sharpiro test are in agreement that the data is not normally distributed. Let's see if log (x +1) transformation can help resolve this issue.

@@

I'll try the Box-Cox transformation first. A limitation of this is that Box Cox cannot values of zero (or negative numbers). We will have to subset observations greater than zero to continue with the transformation. We will use boxcox() to find lambda. 
`library(MASS)`
```{r}
sex_df_DR <- sex_df %>% 
  filter(`Death Rate (estimated)` > 0)
b_DR <-  boxcox(lm(sex_df_DR$`Death Rate (estimated)` ~ 1))
lambda_DR <- b_DR$x[which.max(b_DR$y)]

@@

We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`.We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` to compute the transformation for each observation. Code for `yeo_johnson` obtained from [GitHub](https://stats.stackexchange.com/questions/1444/how-should-i-transform-non-negative-data-including-zeros)
```{r}
library(bestNormalize)

@@

Plotting correlation matrix
```{r message=FALSE, warning=FALSE}
X <- sex_df[,c(9:10, 16:25)]
ggpairs(X)
```
Let's save the dataframe for ROE since this one was able to be normalized after Box Cox transformation.
```{r}
saveRDS(sex_df_ROE, here("data/processed_data/sex_new.rds"))
```

>>>>

Visual assessment and the shapiro test are in agreement that the data is not normally distributed. Let's see if log (x +1) transformation can help resolve this issue.

@@

I'll try the Box-Cox transformation first. A limitation of this is that Box Cox cannot have values of zero (or negative numbers). We will have to subset observations greater than zero to continue with the transformation. We will use boxcox() to find lambda. 
`library(MASS)`
```{r}
sex_df_sub_zero <- sex_df %>% 
  filter(`Death Rate (estimated)` <= 0)

sex_df_DR <- sex_df %>% 
  filter(`Death Rate (estimated)` > 0)
b_DR <-  boxcox(lm(sex_df_DR$`Death Rate (estimated)` ~ 1))
lambda_DR <- b_DR$x[which.max(b_DR$y)]

@@

We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`. We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` to compute the transformation for each observation. Code for `yeo_johnson` obtained from [GitHub](https://stats.stackexchange.com/questions/1444/how-should-i-transform-non-negative-data-including-zeros)
```{r}
library(bestNormalize)

@@

Plotting correlation matrix
```{r message=FALSE, warning=FALSE}
X <- sex_df[,c(9:10, 16:25)]
ggpairs(X)
```

Let's merge back the observations we had to exclude due to limitations of the Box Cox transformation... these will be necessary to carry out spatial analyses..
```{r}
sex_df_ROE <- full_join(sex_df_ROE,sex_df_sub_zero)

Let's save the dataframe for ROE since this one was able to be normalized after Box Cox transformation. 
```{r}
saveRDS(sex_df_ROE, here("data/processed_data/sex_new.rds"))
```

### 2023-04-08 code/processing_code/2a_cleaning_sex_df.Rmd
added more to project

Let's find the HIV death rate by sex. We use round to get x to the closest whole number.
```{r}
sex_df <- sex_df %>% 
  mutate(Male_Rate = round(`Male Death Cases`/pop_male*1e5),
         Female_Rate = round(`Female Death Cases`/pop_female*1e5)) %>% 
  select(!`Year`)
```

>>>>

Let's find the HIV death rate by sex. We use round to get x to the closest whole number.
```{r}
sex_df <- sex_df %>% 
  mutate(Male_Rate = round(`Male Death Cases`/pop_male*1e5),
         Female_Rate = round(`Female Death Cases`/pop_female*1e5))
```

### 2023-04-08 code/processing_code/3_cleaning_us_nabs.Rmd
added more to project

This is the list of neighboring states that was obtained from a user on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state). After some 'investigation' (troubleshooting code), this file is not exactly what I need to create the neighbor list. What is achieved by the following code took me a few seconds to do in Excel with copy and paste. But for the sake of reproducibility, I decided it's well-worth it some spend a little bit of time do this in R.

>>>>

This is the list of neighboring states that was obtained from a user on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state). After some 'investigation' (troubleshooting code), this file is not exactly what I need to create the neighbor list. What is achieved by the following code took me a few seconds to do in Excel with copy and paste... but for the sake of reproducibility, I decided it's well-worth it to some spend a little bit of time do this in R. I'm essentially replicating the dataframe, but crossing over the values on the replicated rows such that both variables contain the same information.

### 2023-04-08 products/manuscript/MANU.qmd
added more to project

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

@@

Question: Can the HIV death rate from 2020 in the United States be modeled by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?) Explain in separate analysis: Race/ ethnicity (Black, Hispanic, White, Multiple) and Sex (Male, Female). Can a spatial model explain variation in HIV death-rates at the state-level and/or regional-level after accounting for variables associated with sociodemographic, SDOH, and HIV-related outcomes?

H0 : There is not a spatial correlation HIV-related deaths and SDOH or HIV-related outcomes.
Ha : There is a spatial correlation between HIV-related deaths and SDOH or HIV-related outcomes.

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

@@

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

@@

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.* Loading the datasets from

Clear the environment and start fresh.

@@

Explore central tendency across demographic groups. Fit a poisson regression model predicting death-to-case ratio from percentage with x SDOH variables and percentage of x demographic group.

>>>>

@@

Question: Can the HIV death rate from 2020 in the United States be modeled by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?) Are these factors correlated with disparities in HIV-related mortality across different racial and ethnic groups? Can a spatial model explain variation in HIV death-rates at the state-level and/or regional-level after accounting for variables associated with sociodemographic, SDOH, and HIV-related outcomes?

H0 : There is not a spatial correlation HIV-related deaths and SDOH or HIV-related outcomes.

Ha : There is a spatial correlation between HIV-related deaths and SDOH or HIV-related outcomes.

@@

I used a number of different methods throughout the various stages of this analysis. To ensure that data importation was simple and reproducible, I used `library(here)`. The vast majority of data cleaning processes were achieved using functions within the `tidyverse`. Some of these processes included renaming, pivoting, selecting, and mutating variables. Additionally, most of the data visualization and maps were obtained using ggplot2, with a few exceptions being packages with functions that are not compatible with the `tidyverse` and simple model diagnostics that base R. During the data visualization process, it became apparent that the dependent variable would require transformation. Data transformation methods of increasing complexity were performed. Ultimately, a Box Cox transformation was able normalize out dependent variable. A small value was added to each observation to ensure that all were greater than 0. To help assess relationships between variables, I performed a correlation analysis, most using principle component analysis (PCA). Data analysis methods were performed in order of increasing complexity, beginning with simple linear regression, then Poisson regression, negative-binomial regression, and finally a conditional autoregessive model (CAR) to account for spatial dependence. Models were judged for goodness of fit using AIC.

@@

All data used in this analysis were obtained from publicly available sources, as note above.

@@

Spatial dependence assessed using Moran's I.
